What We Eval in the Shadows

### 2021-08-10 paper/main.tex
Add remove boilerplate section

\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
letting users pass string argument which can be parsed and evaluated in custom a
environment. An example is the \code{rARS} function shown below. This function
is defined in the \code{AdapSamp} package that implements sampling algorithms.

@@

\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
uses \eval to automatically generate R bindings to native code. The expression
\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
evaluated with different values of \code{what} to generate wrapper functions
which are assigned to the \code{methods} structure keyed by \code{what}.

@@

\paragraph{Extracting \texttt{...} arguments} A surprising use of \eval is the
extraction of unevaluated text of arguments passed to a \code{...} parameter.
Consider the definition of the \code{NVL} function from the
\code{statnet.common} package. This function returns the first argument that is
non \code{NULL}.

@@

\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
\code{igraph} package shown below. This operator returns the result of
evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
it returns the result of evaluating the second argument (\code{rhs}). This is
similar to the \code{\|\|} operator which performs the logical OR of its
arguments except it treats \code{NULL} as the \code{FALSE} value.

@@

\paragraph{Core function variants} Many packages use \eval to implement variations of the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.

@@

\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation make R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
this.

@@

\paragraph{Easier Maintenance} The combination of \c{match.call} and \eval is
frequently used to make it easier to maintain functions with similar interfaces.
This pattern is very commonly employed by modeling functions. Consider the
example below of the \code{survfit.formula} function from package
\code{survival}.

\begin{figure}[h]
\begin{lstlisting}
survfit.formula <- function(formula, data, ...) {
  Call <- match.call()
  indx <- match(c('formula', 'data', ...), names(Call), nomatch=0)
  temp <- Call[c(1, indx)]
  temp[[1L]] <- quote(stats::model.frame)
  mf <- eval.parent(temp)
  ...
}
\end{lstlisting}
  \caption{survival::survfit.formula}
\end{figure}

This function uses \code{match.call} to extract the expression with which it is
called. Then, it extracts the indices of relevant arguments using the
\code{match} function. Next, it subsets the \code{Call} variable to extract
elements at those indices along with the first element which is the function
name. The function name is then replaced with \code{stats::model.frame}.
Finally, this expression is evaluated. By using \code{...}, the function accepts
arbitrary arguments. It filters the arguments of interest using \code{match}
function. \todo{explain more}

\paragraph{Remove Boilerplate} For example, the \code{PopED::get\_rse} uses it to
assign a number of variables in the current scope:
\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
deferred assignment to fill missing values in a column. Package overture
repeatedly evaluates R expressions capturing any assignments that happen in it
as samples of Markov chains.

\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package. It
mutates the variable \code{s} in the package environment. However, the package
environment in R is locked by default, which prevents bindings from being
mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
flagged by the well-formedness checks. To work around this limitation, the
author passes the expression to unlock the package environment as a string to
\code{parse} which is then evaluated by \code{eval}.

@@

\paragraph{Miscellaneous}

\paragraph{Lookup and indexing}

>>>>

\mypara{Simplify Interfaces} A common use of \eval is to simplify interfaces by
letting users pass string argument which can be parsed and evaluated in custom a
environment. An example is the \code{rARS} function shown below. This function
is defined in the \code{AdapSamp} package that implements sampling algorithms.

@@

\mypara{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
uses \eval to automatically generate R bindings to native code. The expression
\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
evaluated with different values of \code{what} to generate wrapper functions
which are assigned to the \code{methods} structure keyed by \code{what}.

@@

\mypara{Extracting varargs} A surprising use of \eval is the
extraction of unevaluated text of arguments passed to a \code{...} parameter.
Consider the definition of the \code{NVL} function from the
\code{statnet.common} package. This function returns the first argument that is
non \code{NULL}.

@@

\mypara{Conditional Evaluation} Developers frequently use \eval to explicitly
control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
\code{igraph} package shown below. This operator returns the result of
evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
it returns the result of evaluating the second argument (\code{rhs}). This is
similar to the \code{\|\|} operator which performs the logical OR of its
arguments except it treats \code{NULL} as the \code{FALSE} value.

@@

\mypara{Core function variants} Many packages use \eval to implement variations of the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.

@@

\mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation make R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
this.

@@

\mypara{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
is frequently used by modeling functions to avoid boilerplate code and ease
maintenance. Consider the example shown below of function \code{survfit.formula}
from the \code{survival} package.

\begin{figure}[h]
\begin{lstlisting}
survfit.formula <- function(formula, data, weights, ...) {
  Call <- match.call()
  indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
  temp <- Call[c(1, indx)]
  temp[[1L]] <- quote(stats::model.frame)
  mf <- eval.parent(temp)
}
\end{lstlisting}
  \caption{survival::survfit.formula}
\end{figure}

Here, \code{match.call} is used to access the expression with which
\code{survfit.formula} is called. The next three lines construct a call to
\code{stats::model.frame} with argument text of parameters \code{'formula'},
\code{'data'}, and \code{'weights'}. Then, the call expression is evaluated in
the parent environment. The use of \code{match.call}, avoids repeated use of
\code{substitute} for extracting the argument expressions. This setup is also
extensible as new parameter names can be easily added to the \code{match} call.

For example, the \code{PopED::get\_rse} uses it to
assign a number of variables in the current scope:
\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
deferred assignment to fill missing values in a column. Package overture
repeatedly evaluates R expressions capturing any assignments that happen in it
as samples of Markov chains.

\mypara{Code Obfuscation} Some packages use \eval to bypass the restrictions
imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. If package authors cannot avoid using these functions, they use \eval
to obfuscate the code, bypassing the checks performed by the tool. An example of
this use is the \code{scalaEnsure} function from the \code{aibd} package. It
mutates the variable \code{s} in the package environment. However, the package
environment in R is locked by default, which prevents bindings from being
mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
flagged by the well-formedness checks. To work around this limitation, the
author passes the expression to unlock the package environment as a string to
\code{parse} which is then evaluated by \code{eval}.

@@

\mypara{Miscellaneous}

\mypara{Lookup and indexing}

\mypara{Attributes}
Package \code{nlme} uses \eval to evaluate expressions attached as attributes to
its objects. Consider the definition of the \code{formula.pdMat} function below.
It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
and evaluates it.

\begin{figure}[h]
\begin{lstlisting}
formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
\end{lstlisting}
  \caption{nlme::formula.pdMat}
\end{figure}
